On Extending a Full-Sharing Multithreaded Tabling Design with Batched Scheduling

### 2015-05-27 paper.tex
paper -> revising 1

Tabling~\cite{Chen-96} is an implementation technique that overcomes
some limitations of traditional Prolog systems in dealing with
redundant sub-computations and recursion. Tabling consists in storing
intermediate answers for subgoals in a proper data structure, called
the \emph{table space}, so that they can be reused when a repeated
subgoal appears during the resolution process. Tabling has become a
popular and successful technique thanks to the ground-breaking work in
the XSB Prolog system and in particular in the SLG-WAM
engine~\cite{Sagonas-98}, the most successful engine of XSB. The
success of SLG-WAM led to several alternative implementations that
differ in the execution rule, in the data-structures used to implement
tabling, and in the changes to the underlying Prolog
engine. Implementations of tabling are now widely available in systems
like Yap Prolog, B-Prolog, ALS-Prolog, Mercury, Ciao Prolog and more
recently Picat.

Multithreading in Prolog is the ability to concurrently perform
computations, in which each computation runs independently but shares
the program clauses. When multithreading is combined with tabling, we
have the best of both worlds, since we can exploit the combination of
higher procedural control with higher declarative semantics. To the
best of our knowledge, XSB~\cite{Marques-08} and Yap~\cite{Areias-12a}
are the only Prolog systems that support the combination of
multithreading with tabling. Yap implements a SWI-Prolog compatible
multithreading library~\cite{Wielemaker-03}. Yap's threads have their
own execution stacks and only share the code area where predicates,
records, flags and other global non-backtrackable data are stored. For
tabled evaluation, a thread views its tables as private but, at the
engine level, Yap has three designs~\cite{Areias-12a} that vary from a
\emph{No-Sharing} (NS) design, where each thread allocates fully
private tables for each new tabled subgoal called during its
computation, to a \emph{Full-Sharing} (FS) design, where threads share
the complete table space.

@@

With the FS design, all tables are shared. Thus, since several threads
can be inserting answers in the same table, when an answer already
exists, it is not possible to determine if the answer is new or
repeated for a certain thread without further support. However, for
local scheduling this is not a problem because, for repeated and new
answers, local scheduling always fails. The problem is with batched
scheduling that requires further support to work with the FS design
since, with batched scheduling, only repeated answers should fail. We
have thus to detect when a newly found answer is also a new answer for
the thread at hand in order to ensure that new answers for a given
subgoal call are correctly propagated once and only once to all
thread's repeated calls.

In this work, we propose an extension to the table space data
structures, which we named \emph{Private Answer Chaining (PAC)}, as a
way to keep track, per thread and subgoal call, of the answers that
were already found and propagated to all repeated calls. We discuss in
detail our proposal for extending the FS design with batched
scheduling and we present a performance analysis comparison between
local and batched scheduling. Experimental results show that, despite
the extra PAC data structures required to support batched scheduling
with the FS design, the execution time of the combination is still
quite competitive when comparing against the FS design with local
scheduling.

@@

This section introduces some background needed for the following
sections.

\subsection{Table Space Organization}

The basic idea behind tabling is straightforward: programs are
evaluated by storing answers for tabled subgoals in an appropriate
data space, called the \emph{table space}. Similar calls to tabled
subgoals are not re-evaluated against the program clauses, instead
they are resolved by consuming the answers already stored in their
table entries. During this process, as further new answers are found,
they are stored in their tables and later returned to all similar
calls.

@@

Local scheduling schedules the evaluation of a program in a
breath-first manner. It favors backtracking first with completion
instead of forward execution, leaving the consumption of answers for
last. Local scheduling only allows a \emph{Cluster of Dependent
  Subgoals} (CDS) to return answers after a fix-point has been
reached~\cite{Freire-96}. In other words, local scheduling tries to
keep a CDS as minimal as possible, thus creating less complex
dependencies between subgoals, which causes a sooner completion of
subgoals.

@@

The algorithm receives two arguments: the newly answer found during
the evaluation (\emph{ANS}) and the subgoal frame which corresponds to
the call at hand (\emph{SF}). The algorithm begins by
checking/inserting the given \emph{ANS} into the answer trie
structure, which will return the leaf node for the path representing
\emph{ANS} (line 1). Then, it checks/inserts the given \emph{leaf}
node into the private chaining for the current thread, which will
return the corresponding answer chain node (line 2). Next in line 3,
it tests whether the answer chain node already existed in the chain,
i.e., if it was inserted or not by the current check/insert operation
in order to return failure (line 4), or it proceeds with marking the
answer \emph{ANS} has found (line 6). At the end (lines 7 to 10), it
returns failure if local scheduling is active (line 8), otherwise,
batched scheduling is active, and it proceeds by propagating the
answer \emph{ANS} to the current execution environment (line 10).

@@

Table~\ref{tab_batched_overhead} shows the overhead ratios, when
compared with the NS design with 1 thread (running with local
scheduling without TabMalloc), for the NS and FS designs, when running
1, 8, 16, 24 and 32 threads, with TabMalloc and, local scheduling
(column \emph{Local}) and batched scheduling (column \emph{Batched})
strategies. In order to give a fair weight to each benchmark, the
overhead ratio is calculated as follows. We begin by running ten times
each benchmark $B$ for each design $D$ with $T$ threads. Then, we
calculate the average of those ten runs and use that value ($D_{BT}$)
to put it in perspective against the base time, which is the average
of the ten runs of the NS design with $1$ thread ($NS_{B1}$). For
that, we use the following formula for the overhead $O_{DBT} = D_{BT}
/ NS_{B1}$. After calculating all the overheads $O_{DBT}$ for a
certain design $D$ and number of threads $T$ corresponding to the
several benchmarks $B$, we calculate the respective minimum, average,
maximum and standard deviation overhead ratios.

>>>>

Tabling~\cite{Chen-96} is a technique that overcomes some limitations
of traditional Prolog systems in dealing with redundant
sub-computations and recursion. Tabling consists in storing
intermediate answers for subgoals in a proper data structure, called
the \emph{table space}, so that they can be reused when a repeated
subgoal appears during the resolution process. Tabling has become a
popular and successful technique thanks to the ground-breaking work in
the XSB Prolog system and in particular in the SLG-WAM
engine~\cite{Sagonas-98}, the most successful engine of XSB. The
success of SLG-WAM led to several alternative implementations that
differ in the execution rule, in the data-structures used to implement
tabling, and in the changes to the underlying Prolog
engine. Implementations of tabling are now widely available in systems
like Yap Prolog, B-Prolog, ALS-Prolog, Mercury, Ciao Prolog and more
recently Picat.

Multithreading in Prolog is the ability to concurrently perform
computations, in which each computation runs independently but shares
the program clauses. When multithreading is combined with tabling, we
have the best of both worlds, since we can exploit the combination of
higher procedural control with higher declarative semantics. To the
best of our knowledge, XSB~\cite{Marques-08} and Yap~\cite{Areias-12a}
are the only Prolog systems that support the combination of
multithreading with tabling. Yap implements a SWI-Prolog compatible
multithreading library~\cite{Wielemaker-03}. Yap's threads have their
own execution stacks and only share the code area where predicates,
records, flags and other global non-backtrackable data are stored. For
multithreaded tabled evaluation, a thread views its tables as private
but, at the engine level, Yap has three
designs~\cite{Areias-12a}. These designs vary from a \emph{No-Sharing}
(NS) design, where each thread allocates private tables for each new
tabled subgoal called during its evaluation, to a \emph{Full-Sharing}
(FS) design, where threads share the complete table space.

@@

With the FS design, all tables are shared. Thus, since several threads
can be inserting answers in the same table, when an answer already
exists, it is not possible to determine if the answer is new or
repeated for a particular thread without further support. For local
scheduling this is not a problem because, for repeated and new
answers, local scheduling always fails. The problem is with batched
scheduling that requires that only the repeated answers should
fail. Threads have then to detect during the evaluation whether an
answer is new in their evaluation and must be propagated or an answer
is repeated and the evaluation should fail.

In this work, we propose an extension to the table space data
structures, which we named \emph{Private Answer Chaining (PAC)}, as a
way to keep track, per thread and subgoal call, of the answers that
were already found and already propagated. We discuss in detail our
proposal for extending the FS design with batched scheduling and we
present a performance analysis comparison between local and batched
scheduling. Experimental results show that, despite the extra PAC data
structures required to support batched scheduling with the FS design,
the execution time of the combination is still quite competitive when
comparing against the FS design with local scheduling.

@@

The basic idea behind tabling is straightforward: programs are
evaluated by storing answers for tabled subgoals in an appropriate
data space, called the \emph{table space}. Similar calls\footnote{We
  are considering \emph{variant-based tabling} similarity. Two tabled
  subgoals A and B are variants if they can be made identical by
  \emph{variable} renaming. For example, \emph{p(X,1,Y)} and
  \emph{p(Y,1,Z)} are variants because both can be transformed into
  \emph{p(VAR0,1,VAR1)}.} to tabled subgoals are not re-evaluated
against the program clauses, instead they are resolved by consuming
the answers already stored in their table entries. During this
process, as further new answers are found, they are stored in their
tables and later returned to all similar calls.

@@

Local scheduling evaluates a tabled logic program in a breath-first
manner. It favors backtracking first with completion instead of
forward execution, leaving the consumption of answers for last. Local
scheduling only allows a \emph{Cluster of Dependent Subgoals} (CDS) to
return answers after a fix-point has been reached~\cite{Freire-96}. In
other words, local scheduling tries to keep a CDS as minimal as
possible, thus creating less complex dependencies between subgoals,
which causes a sooner completion of subgoals.

@@

The algorithm receives two arguments: the newly found answer during
the evaluation (\emph{ANS}) and the subgoal frame which corresponds to
the call at hand (\emph{SF}). The algorithm begins by
checking/inserting the given \emph{ANS} into the answer trie
structure, which will return the leaf node for the path representing
\emph{ANS} (line 1). Then, it checks/inserts the given \emph{leaf}
node into the private chaining for the current thread, which will
return the corresponding answer chain node (line 2). Next in line 3,
it tests whether the answer chain node already existed in the chain,
i.e., if it was inserted or not by the current check/insert operation
in order to return failure (line 4), or it proceeds with marking the
answer \emph{ANS} has found (line 6). At the end (lines 7 to 10), it
returns failure if local scheduling is active (line 8), otherwise,
batched scheduling is active, and it proceeds by propagating the
answer \emph{ANS} to the current execution environment (line 10).

@@

Table~\ref{tab_batched_overhead} shows the overhead ratios, when
comparing against the NS design with 1 thread (running with local
scheduling without TabMalloc), for the NS and FS designs, with 1, 8,
16, 24 and 32 threads, using TabMalloc and, local scheduling (column
\emph{Local}) and batched scheduling (column \emph{Batched})
strategies. In order to give a fair weight to each benchmark, the
overhead ratio is calculated as follows. We begin by running ten times
each benchmark $B$ for each design $D$ with $T$ threads. Then, we
calculate the average of those ten runs and use that value ($D_{BT}$)
to put it in perspective against the base time, which is the average
of the ten runs of the NS design with $1$ thread ($NS_{B1}$). For
that, we use the following formula for the overhead $O_{DBT} = D_{BT}
/ NS_{B1}$. After calculating all the overheads $O_{DBT}$ for a
certain design $D$ and number of threads $T$ corresponding to the
several benchmarks $B$, we calculate the respective minimum, average,
maximum and standard deviation overhead ratios.
